---
title: "exercises_SSD"
author: "Camila Barragan and Ulrich Losener"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( cache = FALSE, comment = NA, message = FALSE, warning = FALSE )
```

## Settings

Required libraries:

```{r}
library(devtools)   # Installing package
library(lme4)   # Multilevel model
library(bain)   # Bayes factor

```

## Installing the Package

```{r}
devtools::install_github("ulrichlosener/BayesSSD")
library(BayesSSD)
```

## Longitudinal Interventions

### Survival function

The sample size determination is designed as two-step process. It is advised to
first determine the survival function, which defines the expected drop-out
pattern of participants. This can be done using the Shiny app
[Attrition](https://losener.shinyapps.io/attrition-survival/​)

#### Additional Exercises

### Sample Size Determination

The second step is to include the attrition function and relevant information
for the sample size:

```{r}
SSD_longit()
```

## Cluster Randomised Trials

For two-arm Cluster Randomised Trial it is important to define beforehand the
hypotheses of study.

-   Pair 1: Null vs. alternative

$H_0: \mu_C = \mu_T$

$H_1:\mu_C<\mu_T$

-   Pair 2: Informative hypotheses

$H_1:\mu_C<\mu_T$

$H_2: \mu_C> \mu_T$

```{r}
# Pair 1 
SSD_crt_null(eff_size = 0.2, n1 = 20, ndatasets = 100, rho = 0.1, BF_thresh1 = 3, BF_thresh0 = 3, eta1 = 0.7, eta0 = 0.7, fixed = "n1", b_fract = 2, max = 200, batch_size = 50)
```

The output includes a table. The second column contains the number of clusters
whilst the third column contains the cluster size. The fourth and fifth columns
contain the empirical power when the null and alternative hypotheses are true,
respectively. The output includes a sensitivity analysis using different values
for fraction b (first column).

```{r}
# Pair 2
SSD_crt_inform(eff_size = 0.5, n1 = 20, ndatasets = 100, rho = 0.1, BF_thresh = 3, eta = 0.8, fixed = "n1")
```

The output include the hypotheses under consideration, the cluster size and
number, the number of clusters, and the empirical power.

In addition to the package, the Shiny app [SSD for CRT with Bayes
factor](https://utrecht-university.shinyapps.io/BayesSamplSizeDet-CRT/) can be
used to determine the sample size for a set of common values of effect size,
ICC, number of clusters, and cluster sizes.

### Additional exercises

**Situation 1 A**

A research teams is designing a study to evaluate the effectiveness of a smoking
prevention programme for children combining the current intervention at school
with an additional intervention at home. The researchers decided to assign
randomly schools to one of the intervention conditions

-   School-Only (control): Prevention curriculum delivered during school hours.

-    School + Home: The standard curriculum plus an integrated home-based family
    component.

According to previous literature about similar interventions, the researchers
expect a Cohen’s d of 0.2 and an ICC of 0.1. A total of 60 schools have agreed
to participate in the study. The researchers decide to aim for a Bayes factor of
at least 5 to provide an evidence they consider substantial. Given the values
mentioned, determine the required number of students per school necessary.

**Situation 1 B**

A research teams is designing a study to evaluate which smoking prevention
programme for children is better. The researchers decided to assign randomly
schools to one of the intervention conditions

-   School-Only: Prevention curriculum delivered during school hours.

-   School + Home: The standard curriculum plus an integrated home-based family
    component.

According to previous literature about similar interventions, the researchers
expect a Cohen’s d of 0.5 and an ICC of 0.025. A total of 60 schools have agreed
to participate in the study. The researchers decide to aim for a Bayes factor of
at least 5 to provide an evidence they consider substantial. Given the values
mentioned, determine the required number of students per school necessary.

**Situation 2 A**

A group of researcher have developed a new programme called FUNBALL to prevent
injuries among football players. In the study, teams assigned to the
intervention condition will complete FUNBALL sessions after their usual warm-up,
while the control condition will maintain their usual training routine.

Based on training data and previous literature, the research group expects an
ICC of 0.025 and an effect size of 0.8. Preliminary recruitment efforts indicate
that 40 players per team are willing to participate. Using the Bayesian
hypothesis testing, the researchers seek a Bayes factor of at least 2. Given
this information, determine the total number of teams that must be recruited for
the study.

**Situation 2 B**

A group of researcher have developed a new programme called FUNBALL to prevent
injuries among football players. To evaluate its efficacy, the researchers have
designed a cluster randomised trial comparing the FUNBALL program to a standard
training control. In the study, teams assigned to the FUNBALL condition will
complete FUNBALL sessions, while the other condition will complete the well
stablished FIFA 11+ programme.

Based on training data and previous literature, the research group expects an
ICC of 0.05 and an effect size of 0.5. Preliminary recruitment efforts indicate
that 5 players per team are willing to participate. Using the Bayesian
hypothesis testing, the researchers seek a Bayes factor of at least 5. Given
this information, determine the total number of teams that must be recruited for
the study.

**Situation 3 A**

A new programme involving Health visitor training in the clinical assessment of
postnatal women is being evaluated. The group of researchers compares the
effectivity of the new programme with the standard  postnatal care. In the
cluster randomised trial, the researchers will allocate randomly the health
centres into the intervention and the control conditions.

Based on previous clinical trials in maternal mental health, the researchers
expect an ICC of 0.1 and an effect size (Cohen’s d) of 0.8. Due to the average
number of mothers at the centres, the cluster size is fixed at 40. The desired
Bayes factor is at least 3. Considering this information, determine the number
of clusters required.

**Situation 3 B**

A clinical research team is investigating which psychological framework is more
effective for treating postnatal depression: a Cognitive Behavioural Approach
(CBA) or a Person-Centred Approach (PCA). In the cluster randomized trial, the
health centres are assigned to one of the two approaches.

Based on previous clinical trials in maternal mental health, the researchers
expect an ICC of 0.1 and an effect size (Cohen’s d) of 0.8. Due to the average
number of mothers at the centres, the cluster size is fixed at 30. The desired
Bayes factor is at least 3. Considering this information, determine the number
of clusters required.

**Based on**:

[Efficacy of a new injury prevention programme (FUNBALL) in young male football
(soccer) players: a cluster-randomised controlled trial -
PubMed](https://pubmed.ncbi.nlm.nih.gov/38499320/)

[Psychological interventions for postnatal depression: cluster randomised trial
and economic evaluation. The PONDER trial \| NIHR Journals
Library](https://www.journalslibrary.nihr.ac.uk/hta/HTA13300)

## Multivariate Cluster Randomised Trial

The function is not still available in the package. However, the Shiny app
[Sample Size Determination for Bayesian Hypothesis Testing in Multivariate
Cluster Randomised Trials](https://cnbi.shinyapps.io/SSD-Multivariate-App/) is
available for sample size determination for a set of realistic values. The tests
for multiple outcomes available are: intersection-union test, omnibus test, and
homogeneity of treatment effects.

# References

Barragan I., C. N. , & Moerbeek, M. (2025). Method for Sample Size Determination
for Cluster-Randomized Trials Using the Bayes Factor. *Journal of Educational
and Behavioral Statistics*, *0*(0).
<https://journals.sagepub.com/doi/10.3102/10769986251362404>

Barragan Ibañez CN, Lösener U, Moeteke N, Moerbeek M (2026) Design and analysis
of behavioral intervention studies: A Bayesian approach. PLoS One 21(2):
e0342163. https://doi.org/10.1371/journal.pone.0342163

---
title: "exercises_SSD"
author: "Camila Barragan and Ulrich Losener"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( cache = FALSE, comment = NA, message = FALSE, warning = FALSE )
```

## Settings

Required libraries:

```{r}
library(devtools)   # Installing package
library(lme4)   # Multilevel model
library(bain)   # Bayes factor

```

## Installing the Package

```{r}
devtools::install_github("ulrichlosener/BayesSSD")
library(BayesSSD)
```

## Longitudinal Interventions

### Survival function

The sample size determination is designed as two-step process. It is advised to
first determine the survival function, which defines the expected drop-out
pattern of participants. This can be done using the Shiny app
[Attrition](https://losener.shinyapps.io/attrition-survival/​)

#### Additional Exercises

### Sample Size Determination

The second step is to include the attrition function and relevant information
for the sample size:

```{r}
SSD_longit()
```

## Cluster Randomised Trials

For two-arm Cluster Randomised Trial it is important to define beforehand the
hypotheses of study.

-   Pair 1: Null vs. alternative

$H_0: \mu_C = \mu_T$

$H_1:\mu_C<\mu_T$

-   Pair 2: Informative hypotheses

$H_1:\mu_C<\mu_T$

$H_2: \mu_C> \mu_T$

```{r}
# Pair 1 
SSD_crt_null(eff_size = 0.2, n1 = 20, ndatasets = 100, rho = 0.1, BF_thresh1 = 3, BF_thresh0 = 3, eta1 = 0.7, eta0 = 0.7, fixed = "n1", b_fract = 2, max = 200, batch_size = 50)
```

The output includes a table. The second column contains the number of clusters
whilst the third column contains the cluster size. The fourth and fifth columns
contain the empirical power when the null and alternative hypotheses are true,
respectively. The output includes a sensitivity analysis using different values
for fraction b (first column).

```{r}
# Pair 2
SSD_crt_inform(eff_size = 0.5, n1 = 20, ndatasets = 100, rho = 0.1, BF_thresh = 3, eta = 0.8, fixed = "n1")
```

The output include the hypotheses under consideration, the cluster size and
number, the number of clusters, and the empirical power.

In addition to the package, the Shiny app [SSD for CRT with Bayes
factor](https://utrecht-university.shinyapps.io/BayesSamplSizeDet-CRT/) can be
used to determine the sample size for a set of common values of effect size,
ICC, number of clusters, and cluster sizes.

### Additional exercises

**Situation 1 A**

A research teams is designing a study to evaluate the effectiveness of a smoking
prevention programme for children combining the current intervention at school
with an additional intervention at home. The researchers decided to assign
randomly schools to one of the intervention conditions

-   School-Only (control): Prevention curriculum delivered during school hours.

-    School + Home: The standard curriculum plus an integrated home-based family
    component.

According to previous literature about similar interventions, the researchers
expect a Cohen’s d of 0.2 and an ICC of 0.1. A total of 60 schools have agreed
to participate in the study. The researchers decide to aim for a Bayes factor of
at least 5 to provide an evidence they consider substantial. Given the values
mentioned, determine the required number of students per school necessary.

**Situation 1 B**

A research teams is designing a study to evaluate which smoking prevention
programme for children is better. The researchers decided to assign randomly
schools to one of the intervention conditions

-   School-Only: Prevention curriculum delivered during school hours.

-   School + Home: The standard curriculum plus an integrated home-based family
    component.

According to previous literature about similar interventions, the researchers
expect a Cohen’s d of 0.5 and an ICC of 0.025. A total of 60 schools have agreed
to participate in the study. The researchers decide to aim for a Bayes factor of
at least 5 to provide an evidence they consider substantial. Given the values
mentioned, determine the required number of students per school necessary.

**Situation 2 A**

A group of researcher have developed a new programme called FUNBALL to prevent
injuries among football players. In the study, teams assigned to the
intervention condition will complete FUNBALL sessions after their usual warm-up,
while the control condition will maintain their usual training routine.

Based on training data and previous literature, the research group expects an
ICC of 0.025 and an effect size of 0.8. Preliminary recruitment efforts indicate
that 40 players per team are willing to participate. Using the Bayesian
hypothesis testing, the researchers seek a Bayes factor of at least 2. Given
this information, determine the total number of teams that must be recruited for
the study.

**Situation 2 B**

A group of researcher have developed a new programme called FUNBALL to prevent
injuries among football players. To evaluate its efficacy, the researchers have
designed a cluster randomised trial comparing the FUNBALL program to a standard
training control. In the study, teams assigned to the FUNBALL condition will
complete FUNBALL sessions, while the other condition will complete the well
stablished FIFA 11+ programme.

Based on training data and previous literature, the research group expects an
ICC of 0.05 and an effect size of 0.5. Preliminary recruitment efforts indicate
that 5 players per team are willing to participate. Using the Bayesian
hypothesis testing, the researchers seek a Bayes factor of at least 5. Given
this information, determine the total number of teams that must be recruited for
the study.

**Situation 3 A**

A new programme involving Health visitor training in the clinical assessment of
postnatal women is being evaluated. The group of researchers compares the
effectivity of the new programme with the standard  postnatal care. In the
cluster randomised trial, the researchers will allocate randomly the health
centres into the intervention and the control conditions.

Based on previous clinical trials in maternal mental health, the researchers
expect an ICC of 0.1 and an effect size (Cohen’s d) of 0.8. Due to the average
number of mothers at the centres, the cluster size is fixed at 40. The desired
Bayes factor is at least 3. Considering this information, determine the number
of clusters required.

**Situation 3 B**

A clinical research team is investigating which psychological framework is more
effective for treating postnatal depression: a Cognitive Behavioural Approach
(CBA) or a Person-Centred Approach (PCA). In the cluster randomized trial, the
health centres are assigned to one of the two approaches.

Based on previous clinical trials in maternal mental health, the researchers
expect an ICC of 0.1 and an effect size (Cohen’s d) of 0.8. Due to the average
number of mothers at the centres, the cluster size is fixed at 30. The desired
Bayes factor is at least 3. Considering this information, determine the number
of clusters required.

**Based on**:

[Efficacy of a new injury prevention programme (FUNBALL) in young male football
(soccer) players: a cluster-randomised controlled trial -
PubMed](https://pubmed.ncbi.nlm.nih.gov/38499320/)

[Psychological interventions for postnatal depression: cluster randomised trial
and economic evaluation. The PONDER trial \| NIHR Journals
Library](https://www.journalslibrary.nihr.ac.uk/hta/HTA13300)

## Multivariate Cluster Randomised Trial

The function is not still available in the package. However, the Shiny app
[Sample Size Determination for Bayesian Hypothesis Testing in Multivariate
Cluster Randomised Trials](https://cnbi.shinyapps.io/SSD-Multivariate-App/) is
available for sample size determination for a set of realistic values. The tests
for multiple outcomes available are: intersection-union test, omnibus test, and
homogeneity of treatment effects.

# References

Barragan I., C. N. , & Moerbeek, M. (2025). Method for Sample Size Determination
for Cluster-Randomized Trials Using the Bayes Factor. *Journal of Educational
and Behavioral Statistics*, *0*(0).
<https://journals.sagepub.com/doi/10.3102/10769986251362404>

Barragan Ibañez CN, Lösener U, Moeteke N, Moerbeek M (2026) Design and analysis
of behavioral intervention studies: A Bayesian approach. PLoS One 21(2):
e0342163. https://doi.org/10.1371/journal.pone.0342163
